---
title: Simple Regression with R
author: Jes LaCourse
date: '2022-10-02'
slug: [SimpleRegression]
categories: ['Regression']
tags: ['Regression', 'plotly']
description: ''
---


# Fitting a Line to Data

Simple linear regression a method of estimating an outcome based on a related variable. In this case, we'll walkthrough estimating systolic blood pressure (our outcome) based on age (a variable) using R.  

Linear regression is presented in point-slope form, with parameters $\beta_0$ and $\beta_1$ making up our slope and intercept, respectively. 

$$y = \beta_0 + \beta_1x$$

We'll want to estimate the systolic blood pressure, $y$, for 20 subjects based on their age, $x$. $\beta$’s are usually unknown, but in this case we know we want to be reasonably close to $\beta_0=90$ and $\beta_1=0.9$. In other words, after many repeated experiments we've determined that the true intercept of our regression line is $90$ mmHg and our true slope is about $0.9$ mmHg/yr.  


$$y = 90 + 0.9x$$



Here are our 20 subjects: 


```{r setup, include = F}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
library(tidyverse)
library(plotly)
library(kableExtra)

theme_set(theme_bw())
```

```{r createSample, echo= F}
# Fix our sample 
set.seed(1002)

# x : I want 20 subjects between 18 and 60
# y : Use our known slope and intercept to set BPs.
x = sample(18:60, 20, replace = T)
y = round(rnorm(20, mean = x*0.9, sd = 15) + 90, 0)  

data <- data.frame(subject = as.character(1:20), age = x, bp = y)
data %>% t %>% kable(table.attr = "style = \"color:white;\"", 
                     caption = "Blood Pressure Sample" ) %>% 
               kable_paper()


```


## First the Basics

It's good to get a feel for the characteristics of our sample. Our subject information is stored in a dataframe called `data`.

We'll want to collect some fundamental information about our sample. 

Call `summary(data)`: 

```{r echo = F}
summary(data) #%>% kable(table.attr = "style = \"color:white;\"", 
              #      ) %>% 
              # kable_paper()
```

The summary tells us we have 20 subjects with a mean age of $35.5$ years. Looking at the quartiles, then our minimum and maximum values, the ages appear to be evenly distributed from 20 to 58 years. A quick check of our blood pressure data shows values that are evenly distributed as well. We'll look at formal diagnostics later, but right now linear regression is looking like a good tool for estimation. 


## Visualizing Data

The function `ggplotly()` from the `plotly` library is a fantastic way to convert `ggplot()` items into interactive figures. Note that you can simply use the `ggplot()` code in the brackets for a static plot.  

```{r, plotPoints}

    {
      ggplot(data, aes(name = subject, x = age, y = bp )) + 
      geom_point() + 
      labs(title = "Blood Pressure by Age, n = 20", 
           y = "Blood Pressure (mmHg)", 
           x = "Age (Years)")
    } %>% 
    ggplotly()
```

We certainly have a linear trend. There also appear to be slightly more younger subjects than older subjects. One of our subjects, `subject 3`, appears to have a blood pressure substantially lower than expected at (56,116). 


## Modeling

```{r modeling}

lm(data, formula = bp ~age) -> m
m
summary(m)
mean(m$residuals^2)

data %>% 
  {
    ggplot(.,aes(age,bp)) +
    geom_point() + 
    geom_line(aes(y = m$coefficients[1] + m$coefficients[2]*x)) +
    geom_line( aes(y = 90+x*0.9), color = "red") +
    theme_bw()
  } %>% 
  ggplotly(., tooltip = list("x", "y"))


```

